Defining properties regulating pioneer-factor binding and function

### 2023-01-23 manuscript/manuscript_sections/_03_Results.qmd
update supplemental figs

-   ~~Description of RNA-seq data: number of differential genes, overlap with ChIP data, GO analysis.~~

-   ~~Comparison of RNA-seq and ATAC data: chromatin opening correlates with transcriptional activation.~~

-   ~~Class III promoters are most likely to be associated with upregulated genes~~

@@

```{r}
twi_ChIP_classes <- here("results/ChIP_peak_classes/twi_ChIP_classes.tsv") |> 
  read_tsv()

@@

Despite the differences between Zld and Grh during Drosophila development, both proteins behaved similarly when expressed in our exogenous S2 cell system.
To provide additional context to our findings, we performed additional experiments with the transcription factor Twist (Twi).
Although Twi is also essential for regulating Drosophila embryonic development, Twi has a number if distnict properties from Zld and Grh, making it an interesting point of comparison.
Twi is a master regulator of mesodermal cell fate (REFS).
Twi has many binding sites that are dynamic during early embryonic development [@Wilczynski2010a], in contrast to Zld and Grh.
Although Twi was hypothesized to be a pioneer factor [@Sandmann2007], pioneering activity by Twi had not previously been tested.
In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012].
To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. XX).
We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
Comparison of Twi ChIP-seq and ATAC-seq data allowed for the identification of the same classes of binding sites that were found for Zld and Grh (Fig 2a,b).
Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
Twi bound extensively to closed chromatin, with `r twi_closed`% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c; Extended Data Fig. 4a).
Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4b).
<!--# Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig. XX). --> GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. 4C).
<!--# Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences. The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig. XX). Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig. XX). This was in contrast to Zld and Grh, where induction results in the activation of many genes that are not directly bound by Zld or Grh. This suggests that Zld and Grh may activate many transcription factors, which go on to activate their own targets, leading to secondary effects in the RNA-seq data. Twi, by contrast, appears to directly activate genes required for mesodermal and muscle cell fate. --> <!--# Is this worth mentioning? Might require additional analysis of how many TFs are upregulated in RNA-seq data for each factor. Could also require additional motif analysis to identify/support secondary TFs and reason for nonspecific effects. --> <!--# Similar to Zld and Grh, transcriptional activation was correlated with chromatin opening (Extended Data Fig. XX). --> Together, our data show that Twi can act as a pioneer factor.

@@

Prior studies have extensively explored the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
To better understand how cell-type specific variables affect pioneer-factor binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
We compared Zld binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig. 7a,b).
Previous work has shown that regions resistant to pioneer-factor binding are enriched for the repressive histone modifications H3K9me3 and H3K27me3 [@Mayran2018a; @Soufi2012].
To test if repressive marks were present at the resistant sites in S2 cells, we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
Based on analysis of these datasets we divided the resistant sites into three additional classes of binding sites: class IV sites have high levels H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a; Extended Data Fig. 8d-f).
We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c; Extended Data Fig. 8e-l).
We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017], and to Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
Despite the presence of class IV, V and VI binding sites for all three proteins, there were distinct patterns of binding across different tissues.
Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig. 4b).
Twi binding sites in S2 cells had low levels of binding in embryos, but many of the strongest embryonic binding sites were unbound.

>>>>

@@

```{r}
#| include: false
twi_ChIP_classes <- here("results/ChIP_peak_classes/twi_ChIP_classes.tsv") |> 
  read_tsv()

@@

Despite the differences between Zld and Grh during Drosophila development, both proteins behaved similarly when expressed in our exogenous S2 cell system.
To provide additional context to our findings, we performed additional experiments with the transcription factor Twist (Twi).
Although Twi is also essential for regulating Drosophila embryonic development, Twi has a number if distnict properties from Zld and Grh, making it an interesting point of comparison.
Twi is a master regulator of mesodermal cell fate (REFS).
Twi has many binding sites that are dynamic during early embryonic development [@Wilczynski2010a], in contrast to Zld and Grh.
Although Twi was hypothesized to be a pioneer factor [@Sandmann2007], pioneering activity by Twi had not previously been tested.
In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012].
To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. 4a).
We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq.
Comparison of Twi ChIP-seq and ATAC-seq data allowed for the identification of the same classes of binding sites that were found for Zld and Grh (Fig 2a,b).
Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c).
Twi bound extensively to closed chromatin, with `r twi_closed`% of binding sites lacking detectible chromatin accessibility (Fig. 2b).
At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c; Extended Data Fig. 4b).
Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. 4c).
<!--# Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig. XX). --> GO analysis demonstrated that genes upregulated upon Twi expression are enriched for mesodermal and muscle-specific genes, consistent with the endogenous function of Twi (Extended Data Fig. 4d).
<!--# Despite the overall similar activities of Zld, Grh, and Twi in S2 cells, we observed several notable differences. The percentage of all closed binding sites that undergo chromatin opening is lower for Twi compared to Zld or Grh (Extended Data Fig. XX). Additionally, almost all of the genes upregulated upon Twi induction are also bound by Twi (Extended Data Fig. XX). This was in contrast to Zld and Grh, where induction results in the activation of many genes that are not directly bound by Zld or Grh. This suggests that Zld and Grh may activate many transcription factors, which go on to activate their own targets, leading to secondary effects in the RNA-seq data. Twi, by contrast, appears to directly activate genes required for mesodermal and muscle cell fate. --> <!--# Is this worth mentioning? Might require additional analysis of how many TFs are upregulated in RNA-seq data for each factor. Could also require additional motif analysis to identify/support secondary TFs and reason for nonspecific effects. --> <!--# Similar to Zld and Grh, transcriptional activation was correlated with chromatin opening (Extended Data Fig. XX). --> Together, our data show that Twi can act as a pioneer factor.

@@

Prior studies have extensively explored the *in vivo* binding sites of Zld, Grh, and Twi during normal Drosophila development.
To better understand how cell-type specific variables affect pioneer-factor binding, we compared ChIP-seq data for our ectopically expressed factors to published datasets for endogenous Zld, Grh and Twi binding.
We compared Zld binding sites in S2 cells to Zld ChIP-seq data from the early embryo ([@Harrison2011], nuclear cycle 14 embryos) and from larval neural stem cells [@Larson2021a].
We identified many binding sites that would normally be bound by Zld in another tissue, but were not bound in S2 cells (Extended Data Fig. 7a,b).
Previous work has shown that regions resistant to pioneer-factor binding are enriched for the repressive histone modifications H3K9me3 and H3K27me3 [@Mayran2018a; @Soufi2012].
To test if repressive marks were present at the resistant sites in S2 cells, we compared our data to published S2 cell ChIP-seq datasets for H3K27me3 [@Jain2020] and H3K9me3 [@Eastwood2021].
Based on analysis of these datasets we divided the resistant sites into three additional classes of binding sites: class IV sites have high levels H3K27me3, class V sites have high levels of H3K9me3, and class VI have low levels of both repressive marks (Fig 4a; Extended Data Fig. 7d-f).
We performed the same comparison for Grh and Twi and identified these same classes (Fig 4b,c; Extended Data Fig. 7e-l).
We compared to Grh ChIP-seq data from late embryos (16-17 hours) and wing imaginal discs [@Nevil2017], and to Twi ChIP-seq data from early embryos (1-3 hours) [@Ozdemir2011].
Despite the presence of class IV, V and VI binding sites for all three proteins, there were distinct patterns of binding across different tissues.
Zld binding in S2 cells was more similar to that of neural stem cells than embryos, with many of the strongest binding sites in embryos remaining unbound in S2 cells (Fig. 4a).
For Grh, by contrast, many of the strongest binding sites were shared across the three tissues (Fig. 4b).
Twi binding sites in S2 cells had low levels of binding in embryos, but many of the strongest embryonic binding sites were unbound.
